Optimal Symmetric Multimodal Templates and Concatenated Random Forests for Supervised Brain Tumor Segmentation (Simplified) with \textit{ANTsR}

### 2014-04-25 paper.tex
ENH:  Addressing Brian's comments.

Instrumental to our success in the 2013 challenge 
was the use of image features based on multimodal shape and appearance 
asymmetries which were generated via the construction of symmetric, 
multivariate brain templates.  Additional improvements in performance
were due to the use of a two-stage RF model approach
whereby the output probability images from the application of the first
RF model seeds the construction of a subset of new feature images
for input into the second stage.  The results of the second RF model
are then refined using a series of heuristically-derived binary morphological
operations.  \textcolor{red}{The two-stage approach is motivated by
the fact that the RF is a powerful tool for ranking the features
in a high-dimensional model i.e. can be used for model selection.
However, the RF outputs are not geometrically constrained, which is
the motivation for our second refinement step.}.

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the approximately 20 teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.
In subsequent sections we describe in greater detail the automated 
supervised brain segmentation pipeline including the generation of asymmetry
feature images and the concatenated RF construction.  We also
describe the code and data%
\footnote{
Although all data was provided having been skull-stripped, 
we note to the reader that an
available ANTs-based solution to the brain extraction problem
is found in the script {\tt antsBrainExtraction.sh} which uses
a hybrid approach based on image registration and binary morphology
to isolate the brain \citep{avants2010a}.  An evaluation with other
methodologies demonstrated comparable to superior performance 
\citep{avants2010a}.
} 
used for the BRATS challenge so that 
the interested reader can reproduce our results.  Finally, we 
illustrate performance with results from the BRATS 2013 challenge.

@@

Certain key elements characterize the supervised brain tumor segmentation
protocol including:
\begin{itemize}
  \item construction of symmetric multivariate templates for multimodal 
        asymmetry-based features,
  \item generation of other image-based features,
  \item training of RF models, and
  \item prediction using the proposed framework.
\end{itemize}
In subsequent sections, we describe these items and detail the
\textit{ANTsR} framework which coordinates all steps.

@@

\begin{figure}[h]
  \centering
  \begin{tabular}{cccc}
  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
  \includegraphics[width=30mm]{Figures/BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
  (a) & (b) &
  (c) & (d) \\
  \end{tabular}
  \caption{Induced bilateral asymmetry due to tumor presence causing
  distortion of the plane of symmetry.  Shown are mid-axial slices of
  one of the BRATS 2013 training data 
  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
  (d) T2.  }
  \label{fig:asymmetry}
\end{figure}

In order to better characterize deviations from normal brain shape 
and appearance, several image features were derived using symmetric 
population-specific multivariate templates.  
For normal neuroanatomy, the use of spatial prior information 
coupled with image normalization capabilities has proven useful 
in producing improved segmentation results of ``expected'' brain tissue
such as cerebrospinal fluid, gray matter, and white matter 
\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
priors to model the presence of a possible tumor and its constituent tissue 
components is difficult. However, since the normal brain 
exhibits a bilaterally symmetric organization, we can 
use the presence of asymmetries to potentially differentiate abnormal 
brain tissue.  A similar motivation prompted
the identification of the mid-sagittal plane of symmetry \citep{prima2002}
for feature generation in multiple sclerosis lesion \citep{geremia2011} and 
tumor \citep{geremia2012} identification.  However, this
approach does not take into account the displacement of 
normal tissue due to tumor growth causing the mid-sagittal plane
to deform from its planar structure (cf Figure \ref{fig:asymmetry}).

\begin{figure}[h]
    \centerline{\includegraphics[width=130mm]{Figures/templateKirby.pdf}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection 
           perpendicular to the mid-sagittal
           plane, $T_R$, resulted in the contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure}

@@

\begin{figure}[!h]
  \centering
    \includegraphics[width=100mm]{Figures/asymmetryFeatureImages.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  Feature images used are outlined in a dashed line.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

As detailed in \cite{avants2008,avants2010}, 
given $K$ multimodality images, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
for $N$ subjects,  multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ to minimize the
following cost function \textcolor{red}{not sure the 2nd term is correct...}:
\begin{align}
  \sum_{n=1}^N 
        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)  
        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically in four iterations), we symmetrize
the template by flipping each asymmetric template component contralaterally 
and then running the
multivariate template construction a second time using only the multivariate
template and its symmetric analog.  This is illustrated conceptually in
Figure \ref{fig:symmetrictemplates}.

In terms of implementation, this template building algorithm is 
encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
available in the ANTs repository, which permits serial or parallel processing on
an individual workstation or on a computational cluster.  In
Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
the MMRR multivariate symmetrical template consisting of FA%
\footnote{
Although DWI-based images were not included in the
challenge data, such images have shown discriminative potential 
\citep{price2003,cha2005} warranting investigation in our future
work.  
}, FLAIR,  
T1, and T2 components.  These images have been made publicly available
along with other processed images such as corresponding
tissue priors and structural labels.%
\footnote{
http://figshare.com/articles/Kirby\_multivariate\_template/852989.
All 7 components are stored under the more informal ``Kirby'' data set moniker.
}

@@

@@

\begin{figure}[!h]
  \centering
  \begin{tabular}{c}
    \includegraphics[width=100mm]{Figures/pipeline2.pdf}
  \end{tabular}
  \caption{Diagrammatic workflow for the proposed RF model training and prediction.  The feature
  images are first generated from the input MRI and symmetric multivariate template. If training,
  the set of feature images derived from the training data are used to create the RF model
  for the first stage.  The model is then applied to the training or prediction data to generate 
  a second set of feature images to create the second stage RF model (if training) and/or generate
  the second RF model.  
  For prediction, the final refinement process entails a series of heuristically-derived
  binary morphological resulting in the final labeled brain.  
  }
  \label{fig:pipeline}
\end{figure}

@@

\textcolor{red}{Nick, I think the motivation for this is still unclear
  .... doesnt it have something to do with the MRF of atropos
  smoothing out the importance features resulting from the RF step?
  Figure 4 tells us a little bit about the procedure of the code but
  the idea is not conveyed ... would it be possible to show exactly
  what is happening in a separate figure?  e.g. just conceptually in a small 
  region of voxels with 2 modalities ....?}

For both stages, in contrast to previous generative
modeling approaches for multimodal tumor segmentation 
\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the RF construction 
process determine the optimal combination of such multivariate
information.  Additionally, maximum posterior labeling from both stages
is used to determine the connected components for each label.  
Geometric features (assigned voxel-wise) include the physical volumes 
of each connected component, the volume to surface area ratio, 
the elongation, and eccentricity.

\paragraph{Stage 1 (GMM)} \textcolor{red}{more descriptive name}
Gaussian mixture modeling (GMM) is used to assign voxelwise
tissue probabilities which comprise an additional set of image features 
where 
prior cluster centers for specific tissue types for each
modality are learned from training data \citep{reynolds2009}.  
Intensity normalization
across the training cohort was originally performed using 
\cite{nyul2000}.  However, despite good performance on normal
data, application of this intensity normalization algorithm 
(after a thorough parameter search) on BRATS data led to slightly 
worse performance than simply winsorizing the intensity values to the quantile
range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
$[0,1]$ so we opted for the latter, more simplistic intensity
normalization protocol.  The cluster centers are defined as the mean
normalized intensity value for each tissue type of each modality 
image over all the training data.  It should be noted that 
initializing with these values is optional and that results
seem to be robust to initialization values.

@@

\textcolor{red}{still feel unclear about it ...  does it go like this?
\begin{enumerate}
\item use tissue prior intensity models (not spatial) to initialize
  Atropos multivariate EM-MRF segmentation of new image  w/o spatial
  priors 
\item get out segmentation and posteriors ... run some shape stats on
  the segmentation 
\item separately compute the asymmetry features by mapping to the MV
  template 
\item input  asymmFeatures + atroposPosteriors + shapeFeatures to the
  RF model  ... with or w/o ground truth outcomes ?  same either way?
  what do you get out of the RF?
\end{enumerate}
}

\paragraph{Stage 2 (MAP-MRF)}

Application of the Stage 1 RF model results in a set of
probabilistic estimates for each of the seven tissue types.  
\textcolor{red}{can you say more here --- what are these in RF and R
  terms?  how the RF generate probabilities ?}
At the beginning of the second stage, these spatial probability maps are 
used as spatial priors in a 
MAP-MRF segmentation also performed using the Atropos 
segmentation tool \citep{avants2011} available in ANTs although we couple
it with N4 bias correction \citep{tustison2010} as contained in the script
{\tt antsAtroposN4.sh}.  The probabilistic segmentation output and corresponding
connected-component feature images for each modality replace the analogous 
GMM-based feature images used 
during the first stage of training/prediction along with the other feature
images generated during the feature stage (such as the template-based data).
The second stage RF model is then produced similarly as the previous stage: 
\begin{align}
\label{eq:mapmrf}
\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP-MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i.
\end{align}
We discovered that this composite application of random
forest models improves the resulting labeled output.
\textcolor{red}{Is this so much a discovery or is it that the MRF
  features ``fix'' the RF output by introducing spatial coherence to
  the results .... at least , that's what it seems to me that you are
  doing ...}

@@

\begin{figure*}[h]
%  \centering
%  \makebox[0.9\textwidth][c]{
  \includegraphics[width=150mm]{Figures/featureImages.pdf}
%  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

\subsection{\textit{ANTsR}:  An ANTs/R Interface}
We briefly describe our imaging-specific interface between
\textit{ANTs} and \textit{R}.

\subsubsection{Overview}

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=120mm]{Figures/challengeResults.pdf}
  \caption{Visual results (every tenth slice sampled from the axial center of the
  tumor region) from the BRATS 2013 challenge using the proposed
  framework (off-white = non-enhancing tumor, cyan = edema, violet = enhancing 
  tumor, yellow = necrosis).  Tumor components were combined into
  three regions for performance assessment ((1) complete tumor: all four tissue
  classes; (2) tumor core:  necrosis, non-enhancing tumor, and enhancing tumor;
  and (3) enhancing tumor).  See Table \ref{table:results} for quantitative measures
  and challenge rankings.
  }
  \label{fig:challengeresults}
\end{figure}

@@

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{Figures/BRATS_HG_GMM.pdf} &
  \includegraphics[width=90mm]{Figures/BRATS_HG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the high-grade glioma
  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy. 
  Features are listed from most to least important.
  }
  \label{fig:hgimportance}
\end{figure*}

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{Figures/BRATS_LG_GMM.pdf} &
  \includegraphics[width=90mm]{Figures/BRATS_LG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the low-grade glioma
  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy.
  Features are listed from most to least important.
  }
  \label{fig:lgimportance}
\end{figure*}

@@

%\section*{Acknowledgments}
\clearpage
\section*{References}

>>>>

Instrumental to our success in the 2013 challenge 
was the use of image features based on multimodal shape and appearance 
asymmetries which were generated via the construction of symmetric, 
multivariate brain templates.  Additional improvements in performance
were due to the use of a two-stage RF model approach
whereby the output probability images from the application of the first
RF model seeds the construction of a subset of new feature images
for input into the second stage.  The results of the second RF model
are then refined using a series of heuristically-derived binary morphological
operations.  The two-stage approach is motivated by
the fact that the RF is a powerful tool for ranking the features
in a high-dimensional model i.e. can be used for model selection.
However, the RF outputs are not geometrically constrained, which is
the motivation for our second refinement step.

In line with our philosophy regarding open science \citep{tustison2013,ince2012}, 
we have made the code available as open source%
\footnote{
https://github.com/ntustison/BRATS2013
}
and, to our knowledge, we are the only competitors out of the approximately 20 teams 
that participated in either years' challenges to do this.  All code is based
on the well-vetted and open source Insight Toolkit (ITK)%
\footnote{
http://www.itk.org/
} 
of the National Institutes of Health and is available through our \textit{ANTsR} 
package.
In subsequent sections we describe in greater detail the automated 
supervised brain segmentation pipeline including the generation of asymmetry
feature images and the concatenated RF construction.  We also
describe the code and data%
%\footnote{
%Although all data was provided having been skull-stripped, 
%we note to the reader that an
%available ANTs-based solution to the brain extraction problem
%is found in the script {\tt antsBrainExtraction.sh} which uses
%a hybrid approach based on image registration and binary morphology
%to isolate the brain \citep{avants2010a}.  An evaluation with other
%methodologies demonstrated comparable to superior performance 
%\citep{avants2010a}.
%} 
used for the BRATS challenge so that 
the interested reader can reproduce our results.  Finally, we 
illustrate performance with results from the BRATS 2013 challenge.

@@

Certain key elements characterize the proposed supervised brain tumor segmentation
protocol including:
\begin{itemize}
  \item construction of symmetric multivariate templates for multimodal 
        asymmetry-based features,
  \item generation of other image-based features,
  \item training of RF models, and
  \item prediction using the proposed framework.
\end{itemize}
In subsequent sections, we describe these items and detail the
\textit{ANTsR} framework which coordinates all steps.

@@

In order to better characterize deviations from normal brain shape 
and appearance, several image features were derived using symmetric 
population-specific multivariate templates.  
For normal neuroanatomy, the use of spatial prior information 
coupled with image normalization capabilities has proven useful 
in producing improved segmentation results of ``expected'' brain tissue
such as cerebrospinal fluid, gray matter, and white matter 
\citep[e.g.,][]{ashburner1997}.  In contrast, accommodating spatial 
priors to model the presence of a possible tumor and its constituent tissue 
components is difficult. However, since the normal brain 
exhibits a bilaterally symmetric organization, we can 
use the presence of asymmetries to potentially differentiate abnormal 
brain tissue.  A similar motivation prompted
the identification of the mid-sagittal plane of symmetry \citep{prima2002}
for feature generation in multiple sclerosis lesion \citep{geremia2011} and 
tumor \citep{geremia2012} identification.  However, this earlier
approach does not take into account the displacement of 
normal tissue due to tumor growth causing the mid-sagittal plane
to deform from its planar structure (cf Figure \ref{fig:asymmetry}).

\begin{figure}
  \centering
  \begin{tabular}{cccc}
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
  \includegraphics[width=30mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_FLAIR_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T1C_slice93.png} &
%  \includegraphics[width=20mm]{BRATS_HG0001/BRATS_HG0001_T2_slice93.png} \\
  (a) & (b) &
  (c) & (d) \\
  \end{tabular}
  \caption{Induced bilateral asymmetry due to tumor presence causing
  distortion of the plane of symmetry.  Shown are mid-axial slices of
  one of the BRATS 2013 training data 
  (specifically {\tt BRATS\_HG0001}):  (a) FLAIR, (b) T1, (c) T1C, and
  (d) T2.  }
  \label{fig:asymmetry}
\end{figure}

\begin{figure*}
    \centerline{\includegraphics[width=130mm]{templateKirby.pdf}}
  \caption{Multivariate symmetric template created from the MMRR 
           data \citep{landman2011}.  Of the seven modalities 
           comprising the set of study acquisitions, we illustrate the
           (a) FA, (b) FLAIR, (c) MPRAGE, and (d) T2 template components.
           Although DWI-based images were not included in the challenge data, 
           such images have shown discriminative potential \citep{price2003,cha2005} 
           warranting investigation in our future work.  
           The optimal transformation and averaging of the individual 
           subject images result in the asymmetric template represented at 
           the top of the middle column.  A horizontal reflection 
           perpendicular to the mid-sagittal
           plane, $T_R$, resulted in the contralateral counterpart represented
           at the bottom of the second column.  The final template seen on
           the right is a result of repeating the template construction using
           the two asymmetric templates as input.
          }
  \label{fig:symmetrictemplates}
\end{figure*}

@@

\begin{figure}
  \centering
%    \includegraphics[width=80mm]{asymmetryFeatureImages.pdf}
    \includegraphics[width=120mm]{asymmetryFeatureImages.pdf}
  \caption{ Given the mapping between the template, $T$, and subject, $S$, domains ($\phi_b: S  \leftrightarrow \underset{b}{\leftrightsquigarrow} T$), various features can be calculated which demonstrate good discriminative qualities.  Feature images used are outlined in a dashed line.
Top:  Difference images with the symmetric multivariate template are created by warping the template to the subject space and performing a voxelwise subtraction from the original modality image.
Bottom:  Similarly, contralateral difference images are calculated from each modality per subject by generating the non-Euclidean contralateral image via the diffeomorphic transform $\phi_b$.  
          }
  \label{fig:asymmetryFeatures}
\end{figure}

As detailed in \cite{avants2008,avants2010}, 
given $K$ multimodality images, ${\mathbf I} = \{I_1,I_2,\ldots, I_K\}$, 
for $N$ subjects,  multivariate 
template construction iterates between optimizing the set 
of diffeomorphic transforms between the subjects and the 
template, 
$\left\{\left(\phi_1,\phi_1^{-1}\right),\ldots,\left(\phi_N,\phi_N^{-1}\right)\right\}$ 
and constructing the 
optimal multivariate template appearance 
$\mathbf{J}=\{J_1,J_2,\ldots, J_K\}$ (with corresponding 
coordinate system $\psi(\mathbf{x})$) to minimize the
following cost function \textcolor{red}{not sure the 2nd term is correct...}:
\begin{align}
  \label{test}
  \sum_{n=1}^N \Bigg[ &D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right) \nonumber  \\
           &+ \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k   
           \left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\Bigg]
\end{align}
%\begin{align}
%  \sum_{n=1}^N 
%        \left[ D \left( \psi(\mathbf{x}),\phi_1^n(\mathbf{x},1)\right)   
%        + \sum_{k=1}^K \lambda_k \Pi_k \left(I_k^n\left(\phi_n(\mathbf{x},0.5)\right),J_k\left(\phi^{-1}_n(\mathbf{x},0.5)\right)\right)\right]
%\end{align}
$D$ is the diffeomorphic shape distance,
\begin{align}
D\left( \phi( \mathbf{x},0),\phi( \mathbf{x},1)\right) = \int_0^1 \| \nu(\mathbf{x},t)\|_L dt
\end{align}
dependent on the choice of linear operator, $L$, and $\nu$
is the velocity field
\begin{align}
\nu\left( \phi(\mathbf{x},t) \right) = \frac{d\phi(\mathbf{x},t)}{dt},\,\,\, \phi(\mathbf{x},0) = \mathbf{x}.
\end{align}
Each pairwise registration employing the similarity metric $\Pi_k$ can 
be assigned a relative weighting, $\lambda_k$, to weight a particular
modality's influence in the construction process.  Once the multivariate
template has converged (typically in four iterations), we symmetrize
the template by flipping each asymmetric template component contralaterally 
and then running the
multivariate template construction a second time using only the multivariate
template and its symmetric analog.  This is illustrated conceptually in
Figure \ref{fig:symmetrictemplates}.

%In terms of implementation, this template building algorithm is 
%encapsulated in the script \verb#antsMultivariateTemplateConstruction.sh#,
%available in the ANTs repository, which permits serial or parallel processing on
%an individual workstation or on a computational cluster.  In
%Figure \ref{fig:symmetrictemplates} we show mid-axial slices from
%the MMRR multivariate symmetrical template consisting of FA%
%\footnote{
%Although DWI-based images were not included in the
%challenge data, such images have shown discriminative potential 
%\citep{price2003,cha2005} warranting investigation in our future
%work.  
%}, FLAIR,  
%T1, and T2 components.  These images have been made publicly available
%along with other processed images such as corresponding
%tissue priors and structural labels.%
%\footnote{
%http://figshare.com/articles/Kirby\_multivariate\_template/852989.
%All 7 components are stored under the more informal ``Kirby'' data set moniker.
%}

@@

\begin{figure*}
\centering
  \makebox[0.9\textwidth][c]{
  \includegraphics[width=150mm]{featureImages.pdf}
  }
  \caption{Representative feature images derived from the 
           BRATS\_HG0301  ``challenge'' data set.
           Neighborhood statistical images for each modality were 
           generated by calculating a given statistic within
           a specified neighborhood radius.  Also calculated for each modality were feature
           images based on either the GMM or the MAP-MRF segmentation.  For the former, we
           show the probability maps for each of the seven labels which are used as feature
           images.  From the resulting hard segmentation, we calculate various geometric 
           measures per connected component of each of the seven labels.  Similarly, the 
           registration to the symmetric template produces the modality-specific 
           difference images with the
           corresponding symmetric template itself and with respect to the 
           contralateral side.   
           This mapping is also used to produce the log Jacobian image.  Finally, the 
           (T1 - TC) image is calculated and, from the
           cerebral mask, we calculate the normalized distance image.  
           }
  \label{fig:featureImages}         
\end{figure*}

@@

\begin{figure}
  \centering
  \begin{tabular}{c}
%    \includegraphics[width=88mm]{pipeline2.pdf}
    \includegraphics[width=110mm]{pipeline2.pdf}
  \end{tabular}
  \caption{Diagrammatic workflow for the proposed RF model training and prediction.  The feature
  images are first generated from the input MRI and symmetric multivariate template. If training,
  the set of feature images derived from the training data are used to create the RF model
  for the first stage.  The model is then applied to the training or prediction data to generate 
  a second set of feature images to create the second stage RF model (if training) and/or generate
  the second RF model.  
  For prediction, the final refinement process entails a series of heuristically-derived
  binary morphological resulting in the final labeled brain.  
  }
  \label{fig:pipeline}
\end{figure}

@@

\begin{figure}
  \centering
  \begin{tabular}{c}
    \includegraphics[width=88mm]{gmm_map_mrf.pdf}
  \end{tabular}
  \caption{Visual comparison between the first and second RF stages
  on the evaluation {\tt BRATS\_HG0001} data set.
  Although the GMM intensity modeling does provide certain 
  discriminative benefit (see Figures \ref{fig:hgimportance}
  and \ref{fig:lgimportance}), the use of the RF-derived posteriors 
  coupled with the described MAP-MRF segmentation framework enhances 
  the accuracy of the intensity-based modeling features.    
  }
  \label{fig:stageComparison}
\end{figure}

%\textcolor{red}{Nick, I think the motivation for this is still unclear
%  .... doesnt it have something to do with the MRF of atropos
%  smoothing out the importance features resulting from the RF step?
%  Figure 4 tells us a little bit about the procedure of the code but
%  the idea is not conveyed ... would it be possible to show exactly
%  what is happening in a separate figure?  e.g. just conceptually in a small 
%  region of voxels with 2 modalities ....?}

For both stages, in contrast to previous generative
modeling approaches for multimodal tumor segmentation 
\citep[e.g.,][]{prastawa2003}, we do not use multivariate 
Gaussians to specify tissue probabilities but rather incorporate each
univariate probability map into the feature vector of the training
data.  As pointed out in \cite{menze2010}, multivariate modeling
might obscure the distinct biological information provided by each 
modality.  Instead, we let the RF construction 
process determine the optimal combination of such multivariate
information.  Additionally, maximum posterior labeling from both stages
is used to determine the connected components for each label.  
Geometric features (assigned voxelwise) include the physical volumes 
of each connected component, the volume to surface area ratio, 
the elongation, and eccentricity.

\paragraph{Stage 1 (Initial voxelwise classification)} 
Gaussian mixture modeling (GMM) is used to assign voxelwise
tissue probabilities which comprise an additional set of image features 
where 
prior cluster centers for specific tissue types for each
modality are learned from training data \citep{reynolds2009}.  
Intensity normalization
across the training cohort was originally performed using 
\cite{nyul2000}.  However, despite good performance on normal
data, application of this intensity normalization algorithm 
(after a thorough parameter search) on BRATS data led to slightly 
worse performance than simply winsorizing the intensity values to the quantile
range $[0.01, 0.99]$ and then rescaling the resulting intensities to 
$[0,1]$ so we opted for the latter, more simplistic intensity
normalization protocol.  The cluster centers are defined as the mean
normalized intensity value for each tissue type of each modality 
image over all the training data.  It should be noted that 
initializing with these values is optional and that results
seem to be robust to initialization values.

@@

\paragraph{Stage 2 (Refined spatial coherence classification)}

Application of the Stage 1 RF model to a set of feature images 
from an individual subjects results in a set of
images estimating the spatial probabilities for each 
of the seven tissue types. At each voxel, the corresponding multi-feature 
vector is propagated through each tree of the random forest 
resulting in a single vote/classification per tree.  The
classification probabilities are derived by normalizing 
these votes to $[0,1]$.

At the beginning of the second stage, these spatial probability maps are 
used as spatial priors in a 
MAP-MRF segmentation also performed using the Atropos 
segmentation tool \citep{avants2011} available in ANTs although we couple
it with N4 bias correction \citep{tustison2010} as contained in the script
{\tt antsAtroposN4.sh}.  The probabilistic segmentation output and corresponding
connected-component feature images for each modality replace the analogous 
GMM-based feature images used 
during the first stage of training/prediction along with the other feature
images generated during the feature stage (such as the template-based data).
The second stage RF model is then produced similarly as the previous stage: 
\begin{align}
\label{eq:mapmrf}
\mathrm{Tissue}\,\,\mathrm{classes} \sim \sum_{i \in MAP-MRF} F_i + \sum_{i \in Asym} F_i + \sum_{i \in Misc} F_i.
\end{align}
The use of these MRF features improves the accuracy of the 
intensity-based modeling over the GMM approach by introducing
a spatial coherence constraint to those features in addition to 
providing a better initialization with the RF posteriors from 
the initial estimation stage.

%\textcolor{red}{Is this so much a discovery or is it that the MRF
%  features ``fix'' the RF output by introducing spatial coherence to
%  the results .... at least , that's what it seems to me that you are
%  doing ...}

@@

\subsection{Overview}

In summary, the proposed RF classification approach per subject consists of the 
following steps:

\begin{enumerate}
  \item Calculate stage 1 features based on:  
  \begin{itemize}
    \item GMM intensity modeling,
    \item asymmetry, and
    \item miscellaneous.
  \end{itemize}
  \item Apply stage 1 RF model to the above calculated features resulting in seven 
        probability maps for the seven tissue types.  
  \item Use the probability maps to calculate stage 2 features based on MAP-MRF intensity modeling.
  \item Apply stage 2 RF model to MAP-MRF, asymmetry, and miscellaneous features to 
  get final RF classification. 
\end{enumerate}

%\textcolor{red}{still feel unclear about it ...  does it go like this?
%\begin{enumerate}
%\item use tissue prior intensity models (not spatial) to initialize
%  Atropos multivariate EM-MRF segmentation of new image  w/o spatial
%  priors 
%\item get out segmentation and posteriors ... run some shape stats on
%  the segmentation 
%\item separately compute the asymmetry features by mapping to the MV
%  template 
%\item input  asymmFeatures + atroposPosteriors + shapeFeatures to the
%  RF model  ... with or w/o ground truth outcomes ?  same either way?
%  what do you get out of the RF?
%\end{enumerate}
%}

\subsection{\textit{ANTsR}:  An ANTs/R Interface}

@@

\begin{figure*}
  \vspace{-20mm}
  \makebox[\textwidth][c]{
  \includegraphics[width=140mm]{challengeResults.pdf}
  }
  \caption{Visual results (every tenth slice sampled from the axial center of the
  tumor region) from the BRATS 2013 challenge using the proposed
  framework (off-white = non-enhancing tumor, cyan = edema, violet = enhancing 
  tumor, yellow = necrosis).  Tumor components were combined into
  three regions for performance assessment ((1) complete tumor: all four tissue
  classes; (2) tumor core:  necrosis, non-enhancing tumor, and enhancing tumor;
  and (3) enhancing tumor).  See Table \ref{table:results} for quantitative measures
  and challenge rankings.
  }
  \label{fig:challengeresults}
\end{figure*}

@@

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{BRATS_HG_GMM.pdf} &
  \includegraphics[width=90mm]{BRATS_HG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the high-grade glioma
  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy. 
  }
  \label{fig:hgimportance}
\end{figure*}

\begin{figure*}
  \makebox[\textwidth][c]{
  \begin{tabular}{cc}
  \includegraphics[width=90mm]{BRATS_LG_GMM.pdf} &
  \includegraphics[width=90mm]{BRATS_LG_MAP_MRF.pdf} \\
  \end{tabular}
  }
  \caption{{\tt MeanDecreaseAccuracy} plots generated from the low-grade glioma
  Stage 1 and Stage 2 RF models.  These plots provide a weighted 
  ranking describing the importance of each feature for predictive accuracy.
  }
  \label{fig:lgimportance}
\end{figure*}

@@

%\section*{Acknowledgments}
%\clearpage
\section*{References}
